Efficiently Handling Guards in the Low Level Design of RPython's Tracing JIT

### 2012-08-10 talk/vmil2012/paper.tex
more rewriting of the evaluation section

\bivab{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
    \begin{lstlisting}[mathescape]
$b_1$ = int_eq($i_2$, 1)
guard_false($b_1$)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Separated and merged compilation of operations and guards}
  \label{fig:trace-compiled}
\end{figure}

@@

From the mentioned benchmarks we collected different datasets to evaluate the
Frequency, the overhead and overall behaviour of guards, the results are
summarized in the remainder of this section.

\subsection{Frequency of Guards}
\label{sub:guard_frequency}
\begin{figure*}
    \include{figures/benchmarks_table}
    \caption{Benchmark Results}
    \label{fig:benchmarks}
\end{figure*}
Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. Showing that the overall optimization rate is between
65.80\% and 86.23\% of all operations and that the optimization rate for guards
is similar to the general one, as could be assumed based on
Figure~\ref{fig:guard_percent}. These numbers show that guards are a rather
common operation in the traces, which is a reason the put effort into
optimizing them.
\subsection{Overhead of Guards}
\label{sub:guard_overhead}
\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume Data sizes in KiB}
    \label{fig:resume_data_sizes}
\end{figure}

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The size of the machine code is composed
of the size of the compiled operations, the trampolines generated for the
guards and a set of support functions that are generated when the JIT starts
and are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variable and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data\todo{explain
why it is an approximation}.

Compared to the size of the generated machine code the compressed
\texttt{low-level resume data} is about 15\% to 20\% of that size, depending on
the benchmark. On the other hand the generated machine code has only a size
ranging from 20.21\% to 37.98\% of the size of the high and low-level
\texttt{resume data} being compressed as described before.

Tracing JIT compilers only compile the subset of the code executed in a program
that is traced in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches. Still
the overhead associated to guards to resume execution from a side exit appears
to be high.\bivab{put into relation to other JITs, compilers in general}

Both figures do not take into account garbage collection. Pieces of machine
code can be globally invalidated or just become cold again. In both cases the
generated machine code and the related data is garbage collected. The figures
show the total amount of operations that are evaluated by the JIT and the
total amount of code and data that is generated from the optimized traces.

>>>>

\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
    \begin{lstlisting}[mathescape]
$b_1$ = int_eq($i_2$, 1)
guard_false($b_1$)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Separated and merged compilation of operations and guards}
  \label{fig:trace-compiled}
\end{figure}

@@

From the mentioned benchmarks we collected different datasets to evaluate the
Frequency, the overhead and overall behaviour of guards, the results are
summarized in the remainder of this section. We want to point out three
aspects of guards in particular
\begin{itemize}
  \item Guards are very common operations in traces.
  \item There is overhead associated with guards.
  \item Guard failures are local and rare.
\end{itemize}

All figures in this section do not take into account garbage collection. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

\subsection{Frequency of Guards}
\label{sub:guard_frequency}
\begin{figure*}
    \include{figures/benchmarks_table}
    \caption{Benchmark Results}
    \label{fig:benchmarks}
\end{figure*}

Figure~\ref{fig:benchmarks} summarizes the total number of operations that were
recorded during tracing for each of the benchmarks and what percentage of these
operations are guards. The number of operations was counted on the unoptimized
and optimized traces. The Figure shows that the overall optimization rate for
operations which is between 69.4\% and 83.89\% of the traced operations and the
optimization rate of guards, which is between 65.8\% and 86.2\% of the
operations, are very similar, as could be assumed based on
Figure~\ref{fig:guard_percent}. This indicates that the optimizer can remove
most of the guards, but after the optimization pass guards still account for
15.2\% to 20.2\% of the operations being compiled and later executed, the
frequency of this operation makes it important to store the associated
information efficiently and also to make sure that guard checks are executed
fast.

\subsection{Overhead of Guards}
\label{sub:guard_overhead}
\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume Data sizes in KiB}
    \label{fig:resume_data_sizes}
\end{figure}

The overhead that is incurred by the JIT to manage the \texttt{resume data},
the \texttt{low-level resume data} as well as the generated machine code is
shown in Figure~\ref{fig:backend_data}. It shows the total memory consumption
of the code and of the data generated by the machine code backend and an
approximation of the size of the \texttt{resume data} structures for the
different benchmarks mentioned above. The machine code taken into account is
composed of the compiled operations, the trampolines generated for the guards
and a set of support functions that are generated when the JIT starts and which
are shared by all compiled traces. The size of the \texttt{low-level resume
data} is the size of the compressed mapping from registers and stack to
IR-level variables and finally the size of the \texttt{resume data} is an
approximation of the size of the compressed high-level resume data as described
in Section~\ref{sec:Resume Data}\todo{explain why it is an approximation}.

For the different benchmarks the \texttt{low-level resume data} has a size of
about 15\% to 20\% of the amount of memory compared to the size of the
generated machine code. On the other hand the generated machine code has only a
size ranging from 20.5\% to 37.98\% of the size of the high and low-level
\texttt{resume data} combined and being compressed as described before.

Tracing JIT compilers only compile the subset of the code executed in a program
that is traced in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the \texttt{resume data} when
compared to the illustrates why it is important to compress this information.
